Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest

### 2021-04-09 09.0_manuscript.rmd
Explained uncertainty in isotopic data

Section  | Word Count  
------------- | -------------  
Summary | 200  
Main Text  | 6971  
Introduction | 832  
Methods |  3211  
Results | 855  
Discussion | 2073  
Keywords | 8

@@

\
In this paper, we estimate plant-available water in the whole-soil column in a tropical forest, and inversely estimate rooting depths of co-occurring tree species from their growth responses, with a series of model calibrations and validation. We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates through several census-intervals over a 35 year period that experienced El Niño droughts of a variety of intensity, frequency and duration [@Detto.2018; @Condit.2017]. Our hypotheses are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth (see Tables \@ref(tab:symbols) and \@ref(tab:hypo)); and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower exposure to water-stress compared to shallow-rooted species. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic traits and multi-decadal mortality outcomes for a species-rich tropical forest.

@@

Roots could impact tree growth through two factors: 1) water uptake; and 2) nutrient uptake. Because our study is situated in an old-growth tropical forest, we assume that nutrients are mainly concentrated in the shallow soil layers derived from litter decomposition, and that rooting depths do not substantially affect nutrient uptake. Because our study site has a dry season, we expect that rooting depth is a key factor affecting tree growth. We defined species-specific effective rooting depth (ERD) as the depth at which a soil moisture growth limitation factor, $B$, determined by the soil water potential and hydraulic traits best explained species’ diameter growth dynamics. $B$ is used to approximate the amount of stomatal closure due to water-stress in the soil. In this study, we used the fractional loss of hydraulic conductivity of the leaf ($\textrm{FLC}$~leaf~) to estimate $B$. $\textrm{FLC}$~leaf~ is calculated as a fraction of maximum leaf conductivity $K_\textrm{max, leaf}$:

@@

where $K$~leaf~ is the current leaf hydraulic conductance and $K_\textrm{max, leaf}$ is the maximum leaf hydraulic conductance (see Table \@ref(tab:symbols) for trait symbols and their definitions). The dynamics of $K$~leaf~ are estimated from the species-specific relationship between $K$~leaf~ vs. $\mathrm{\Psi}$~leaf~, referred to as the leaf hydraulic vulnerability curve [@Sack.2012]. The vulnerability to loss of hydraulic conductivity arises not only from embolism but also from extra-xylem processes [@Scoffoni:2017]. Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, this allowed us to substitute $\mathrm{\Psi}$~leaf~ with $\mathrm{\Psi}$~soil~ in leaf hydraulic vulnerability curves for each species $s$---defined below in Eq. \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$---to predict maximum diurnal $K_{\textrm{leaf},s, i}$ for each day $i$ as:

\begin{equation}
K_{\textrm{leaf},s, i} = A_{s}e^{-B_{s}{\mathrm{\Psi}}_{\textrm{soil, i}}}
(\#eq:k-soil)
\end{equation}
$K_{\textrm{max, leaf}, s}$, was obtained as the maximum value of $K_{\textrm{leaf}, s}$ using Eq. \@ref(eq:k-soil). $K$~leaf~ is strongly related to photosynthetic capacity [@Brodribb.2002]. Because we were interested in relating $K$~leaf~ to 5-year average diameter growth observations, we ignored diurnal dynamics of $K$~leaf~.

@@

\
We therefore tested alternate structures of empirical growth models, or inverse models of ERD, in which we regressed species-specific growth against multiplicative or additive effects of one or more growth factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $textrm{FLC}_\textrm{leaf}$ and the leaf area index (LAI) (Eq. \@ref(eq:M-chosen) and those in SI). Incorporation of radiation did not improve model fitting, possibly due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI.

@@

\
We evaluated different model structures (see SI) and for each we estimated species effective rooting depths (ERD) as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of growth via modeled growth (e.g. Eq.\@ref(eq:M-chosen)). Our growth model, or inverse ERD model, does not explicitly use rooting profiles, but identifies soil water dynamics at a single depth $z$ as the central tendency that influences the observed growth dynamics the most. We modelled multiple hydrological realisations of soil water-potential dynamics (see below). Incorporating this uncertainty, we defined species ERD as the median ($\pm$ SE) of best-fit depths across all hydrological realisations. See SI for statistics for identifying best-fit ERD.

@@

\
As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq. \@ref(eq:vc-exp). We thus obtained parameters $A$ and $B$ for all the `r length(erd.sp)` ERD species for use in Eq. \@ref(eq:b-mod). We also estimated species $\mathrm{\Psi}$~20,leaf~ using their vulnerability curves.

@@

In some of the alternative models for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI) on growth. We assumed species-specific mean seasonal curves for LAI (standardised between 0 and 1; unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime (@Osnas.2018; see SI).

@@

To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM, @Caldwell.2019), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES, @Koven:2020) (hereafter, ELM-FATES). ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes. In particular, soil hydrologic fluxes are vertically resolved among discrete soil layers (1-D) similarly to the CLM4.5 [@Oleson.2013]. FATES is a community-based, open-source model used for studying climate-vegetation interactions. FATES is a vegetation demography model, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography [@Moorcroft.2001] approach. FATES couples to ELM by a common interface of water and carbon fluxes. Detailed descriptions for ELM and FATES can be found elsewhere [@Bisht.2018; @Fisher.2010; @Fisher.2014; @Koven:2020].

@@

To parameterize catchment hydrology in ELM-FATES, we identified eleven parameters relevant for the water balance, determined their ranges based on literature for the study site, else for the tropics (Table S1), and ran 5000 simulations using Latin Hypercube Sampling [LHS; @Stein.1987] from this global parameter space. Notably, we leveraged local data for soil hydraulic conductivity by depth (@Godsey:2004; Fig. S2), and instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs. $\mathrm{\Psi}$~soil~ [@Kupers:2019]. See SI for details.

@@

We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in the water balance equation, namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot (2012-2017), (2) local stream discharge (2012-2018; see @Paton.2019) and (3) soil volumetric water content (VWC) from two sources: (i) a long-term (2012-2018) record of VWC averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats (@Kupers:2019).

\
 For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized RMSE between observations and simulations across all fluxes and states mentioned above, and then identified 100 parameter ensemble members that minimised the objective function, ensuring that soil moisture dynamics by depth is captured correctly. See SI for details and Table S1 for the ranges of best-fit values for different parameters.

@@

We ran ELM-FATES with the best-fit 100 ensemble members from 1985-2018, with the first 5 years used for model spin-up. Extreme hydrological droughts were identified by depth $z$ as days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.

@@

We used $\mathrm{\delta}$^2^H~xylem~ as an independent observation to validate the ERD models. As root water-uptake is generally a non-fractionating process, tree $\mathrm{\delta}$^2^H~xylem~ reflects the signature of source water. Given a vertical gradient of $\mathrm{\delta}$^2^H in soil and groundwater $\mathrm{\delta}$^2^H~xylem~ provides an index of rooting depth [@Dawson.1991]. We leveraged $\mathrm{\delta}$^2^H~xylem~ from BCI [@Meinzer:2001] for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species and vertically in soil and groundwater $\mathrm{\delta}$^2^H at natural abundance level (Fig. S3). Analyzing $\mathrm{\delta}$^2^H~xylem~ for 12 tree species @Meinzer:1999 found that four species sourced water deeper than 1 m, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.

\
For comparison with modelled ERD, we removed six species from the @Meinzer:1999 dataset to account for the uncertainty in their water-sourcing depths. One species, *Guapira standleyana*, was removed because it's $\mathrm{\delta}$^2^H value of -33.5 (SE $\pm$ 2.6) was particularly difficult to assign a source-water depth given the uncertainty in soil water $\mathrm{\delta}$^2^H data by depth (see Fig S3). $\mathrm{\delta}$^2^H~xylem~ from leafless trees may not be linked with water sourced at the time of measurement and thus may not be comparable to species that had leaves. Leaflessness status of sampled trees is not recorded in @Meinzer:2001. We therefore removed five species that are typically leafless in March-April (Joseph Wright, personal observation), that is, the months of $\mathrm{\delta}$^2^H~xylem~ sampling. For each model of ERD (Eq. \@ref(eq:M-chosen) and alternatives in SI), modeled species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~ for a maximum of six species.

@@

We introduce a novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil water potential profiles, and observed VPD (\@ref(eq:M-chosen), Figs. \@ref(fig:schematic), \@ref(fig:sp-erd)). Our predictions of ERD were consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Figs. \@ref(fig:erd-iso)). Our analyses suggest that co-occurring large canopy tree species with deeper ERD were associated with higher above-ground hydraulic efficiency, but lower safety (Fig. \@ref(fig:erd-hyd)). Nevertheless, evergreen but not deciduous species with deeper ERD showed significantly lower mortality rates than shallower ERD species (Fig. \@ref(fig:erd-mort-census-evg), S11). This ERD-mortality relationship was significant in five out of six census-intervals that had experienced one or more El-Niño events---over seven census-intervals studied in total (1982-2015). Quantifying extreme droughts in the whole soil column over 1990-2015 revealed that ERD explained significant levels of mortality in periods when extreme soil droughts prolonged the dry season water-stress (Fig. \@ref(fig:swp-droughts)). Species exposure to water-stress exponentially declined with deeper ERD (Fig. \@ref(fig:erd-time-evg)). Because extreme, sustained water-stress increases risk of mortality via hydraulic-failure and/or carbon starvation, deeper ERD may mitigate drought-induced mortality by limiting exposure to water-stress [@Rowland:2015; @McDowell:2018]. We thus demonstrate, for the first time, a link between species trade-offs in above- and below-ground hydraulic traits, drought exposure through extreme hydrological droughts quantified over the whole soil-column across several El-Niño events, and large ($\ge$ 10 cm DBH) tree mortality over 35 years.

@@

\
We found significant ERD-mortality relationship during extreme water-stress for evergreen but not deciduous species (Fig. \@ref(fig:erd-mort-census-evg), S11); consistent with the expectation that deciduous species can also avoid water-stress via leaf-drop. This is also consistent with the observation that species distribution along local (BCI) and regional (Panama) moisture gradients is correlated with $\mathrm{\Psi}$~tlp~ for evergreen but not deciduous species [@Kunert.2021]. Studies assessing coordination between leaf phenology, rooting depth and hydraulic traits simultaneously are almost absent for the tropics and warrant future consideration [@Oliviera.2021]. Our analysis for ERD-hydraulic traits relationship was limited to only seven species, but covered all species of different leaf phonologies (Fig. \@ref(fig:erd-hyd)), so here we assume that the trend in the ERD-hydraulic trait relationships holds true across all species of different leaf phonologies. We also found that the range of ERDs overlapped across leaf phenology, but evergreen species tend to have deeper ERDs (Fig. \@ref(fig:sp-erd)), consistent with observations by @Meinzer:1999 and elsewhere [@Smithmartin.2020; @Fan:2017; @Oliviera.2021]. Whether deciduous species also have more efficient and vulnerable hydraulics at BCI as is observed elsewhere remains to be studied [@Markesteijn:2010; @Markesteijn:2011; @Xu:2016; @Gleason.2016]. Leaf phenologies at BCI are numerous and complicated and warrant research.

@@

\
To estimate topographic variation in ERD within or across species, future studies may use soil moisture dynamics from a distributed hydrological model [e.g. @Schwantes:2018]. We note that data availability on soil water retention curves ($\mathrm{\Psi}$~soil~ vs. VWC), hydraulic conductivity (K~soil~) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model. We recommend widespread and coordinated collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.

@@

\
Figure \@ref(fig:erd-mort-census-evg): Mortality rate (mean $\pm$ 1SE in % yr^-1^) versus modeled effective rooting depth (ERD; mean $\pm$ 1SE in m) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen, canopy species found on Barro Colorado Island over seven census intervals (1981-2015). $R$^2^ and significance levels for linear model fits are given in panel insets. Model fits (blue lines) with confidence bands (gray area) are only shown for periods with significant fits at $\alpha$ = 0.05, identified by census-intervals with a superscript *.

\
Figure \@ref(fig:erd-time-evg): Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen species included in the mortality analyses (Fig. \@ref(fig:erd-mort-evg)). Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ was more negative than species $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~20,leaf~, and where $z$ is the soil depth matching species ERD. Standard error of the mean are shown over each bar when available. Note the squared y-axis scale.

>>>>

Section  | Word Count  
------------- | -------------  
Summary | 200  
Main Text  | 6939  
Introduction | 832  
Methods |  3150  
Results | 855  
Discussion | 2102  
Keywords | 8

@@

\
In this paper, we estimate plant-available water in the whole-soil column in a tropical forest, and inversely estimate rooting depths of co-occurring tree species from their growth responses, with a series of model calibrations and validation. We evaluate how rooting depth is linked to above-ground hydraulic traits and mortality rates through seven census-intervals over a 35-year period that experienced El Niño droughts of a variety of intensity, frequency and duration [@Detto.2018; @Condit.2017]. Our hypotheses are that (i) deep-rooted trees have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem due to greater and more reliable water availability at depth (see Tables \@ref(tab:symbols) and \@ref(tab:hypo)); and that, (ii) deep-rooted species have lower mortality rates during droughts due to a lower exposure to water-stress compared to shallow-rooted species. To our knowledge this is the first study to test for a mechanistic link between plant available water in the whole-soil column, trees' above- and below-ground hydraulic traits and multi-decadal mortality outcomes for a species-rich tropical forest.

@@

Roots could impact tree growth through two factors: 1) water uptake; and 2) nutrient uptake. Because our study is situated in an old-growth tropical forest, we assume that nutrients are mainly concentrated in the shallow soil layers derived from litter decomposition, and that rooting depths do not substantially affect nutrient uptake. Because our study site has a dry season, we expect that rooting depth is a key factor affecting tree growth. We defined species-specific effective rooting depth (ERD) as the depth at which a soil moisture growth limitation factor, $B$, determined by the soil water potential and hydraulic traits best explained species’ diameter growth dynamics. $B$ is used to approximate the amount of stomatal closure due to water-stress in the soil. In this study, we used the fractional loss of hydraulic conductivity of the leaf ($\textrm{FLC}$~leaf~) to estimate $B$. $\textrm{FLC}$~leaf~ is calculated as a fraction of maximum leaf conductivity $K_\textrm{max, leaf}$,

@@

where $K$~leaf~ is the current leaf hydraulic conductance and $K_\textrm{max, leaf}$ is the maximum leaf hydraulic conductance (see Table \@ref(tab:symbols) for trait symbols and their definitions). The dynamics of $K$~leaf~ are estimated from the species-specific relationship between $K$~leaf~ vs. $\mathrm{\Psi}$~leaf~, referred to as the leaf hydraulic vulnerability curve [@Sack.2012]. The vulnerability to loss of hydraulic conductivity arises not only from embolism but also from extra-xylem processes [@Scoffoni:2017]. Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, this allowed us to substitute $\mathrm{\Psi}$~leaf~ with $\mathrm{\Psi}$~soil~ in leaf hydraulic vulnerability curves for each species $s$---defined below in Eq. \@ref(eq:vc-exp) using species-specific parameters $A_{s}$ and $B_{s}$---to predict maximum diurnal $K_{\textrm{leaf},s, i}$ for each day $i$,

\begin{equation}
K_{\textrm{leaf},s, i} = A_{s}e^{-B_{s}{\mathrm{\Psi}}_{\textrm{soil, i}}}
(\#eq:k-soil)
\end{equation}
$K_{\textrm{max, leaf}, s}$ was obtained as the maximum value of $K_{\textrm{leaf}, s}$ using Eq. \@ref(eq:k-soil). $K$~leaf~ is strongly related to photosynthetic capacity [@Brodribb.2002]. Because we were interested in relating $K$~leaf~ to 5-year average diameter growth observations, we ignored diurnal dynamics of $K$~leaf~.

@@

\
We therefore tested alternate structures of empirical growth models (Eq. \@ref(eq:M-chosen) and those in SI), or inverse models of ERD, in which we regressed species-specific growth against multiplicative or additive effects of one or more growth factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $\textrm{FLC}_\textrm{leaf}$ and the leaf area index (LAI). Incorporation of radiation did not improve model fitting, possibly due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI.

@@

\
We evaluated different model structures and for each we estimated species effective rooting depths (ERD) as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of growth $G$ (see below) via modeled growth $\hat{G}$. Our growth model, or inverse ERD model, does not explicitly use rooting profiles, but identifies soil water dynamics at a single depth $z$ as the central tendency that influences the observed growth dynamics the most. We modelled multiple hydrological realisations of soil water-potential dynamics (see below). Incorporating this uncertainty, we defined species ERD as the median ($\pm$ SE) of best-fit depths across all hydrological realisations. See SI for statistics for identifying best-fit ERD.

@@

\
As these fits explained a large proportion of variation in parameters $A$ and $B$ (see results), we sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r sp.n.model` ERD species without direct data and estimated leaf hydraulic vulnerability curves using Eq. \@ref(eq:vc-exp). We thus obtained parameters $A$ and $B$ for all the `r length(erd.sp)` ERD species for use in Eq. \@ref(eq:k-soil). We also estimated species $\mathrm{\Psi}$~20,leaf~ using their vulnerability curves.

@@

In some of the alternative models for effective rooting depth, we explored the effect of seasonality in leaf area index (LAI) on growth. We assumed species-specific mean seasonal curves for LAI (standardised between 0 and 1; unitless), informed by a combination of long-term records for leaf-fall [@Wright.1990aeo] and leaf-lifetime [@Osnas.2018]. See SI.

@@

To estimate dynamics of water availability by depth over the forest’s rooting zone, we used the Energy Exascale Earth System Land Model (ELM; @Caldwell.2019), coupled with the Functionally Assembled Terrestrial Ecosystem Simulator (FATES; @Koven:2020) (hereafter, ELM-FATES). ELM is a land-model that, among many features, simulates the physics and conservative dynamics of water, energy and carbon fluxes. In particular, soil hydrologic fluxes are vertically resolved among discrete soil layers (1-D) similarly to the CLM4.5 [@Oleson.2013]. FATES is a community-based, open-source model used for studying climate-vegetation interactions. FATES is a vegetation demography model, with a size-structured group of plants (cohorts) and successional trajectory-based patches based on the ecosystem demography [@Moorcroft.2001] approach. FATES couples to ELM by a common interface of water and carbon fluxes. Detailed descriptions for ELM and FATES can be found elsewhere [@Bisht.2018; @Fisher.2010; @Fisher.2014; @Koven:2020].

@@

To parameterize catchment hydrology in ELM-FATES, we identified eleven parameters relevant for the water balance, determined their ranges based on literature for the study site, else for the tropics (Table S1), and ran 5000 simulations using Latin Hypercube Sampling [LHS; @Stein.1987] from this global parameter space. Notably, we leveraged local data for soil hydraulic conductivity by depth (@Godsey:2004; Fig. S2), and instead of the ELM default soil texture-based pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content vs. $\mathrm{\Psi}$~soil~ [@Kupers:2019]. See SI.

@@

We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in the water balance equation, namely, (1) evapotranspiration ET from the flux tower by the 50 ha plot (2012-2017), (2) local stream discharge (2012-2018; see @Paton.2019) and (3) soil volumetric water content (VWC) from two sources: (i) a long-term (2012-2018) record of VWC averaged across three vertical time domain reflectometry (TDR) probes over depth 0-15 cm from three locations near the flux tower, and (ii) plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 at depths of 0.15, 0.4 and 1 m (1299 samples covering all soil types and habitats; @Kupers:2019).

\
 For ELM-FATES calibration we calculated an objective function for each of the 5000-member ensemble by equally weighting standardized root mean square error (RMSE) between observations and simulations across all fluxes and states mentioned above, and then identified 100 parameter ensemble members that minimised the objective function, ensuring that soil moisture dynamics by depth is captured correctly (see SI). See Table S1 for the ranges of best-fit values for different parameters.

@@

We ran ELM-FATES with the best-fit 100 ensemble members from 1985-2018, with the first five years used for model spin-up. Extreme hydrological droughts were identified by depth $z$ as days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was more negative than the 5^th^ percentile of $\mathrm{\Psi}_{\textrm{soil},z}$ for a given day of the year.

@@

We used $\mathrm{\delta}$^2^H~xylem~ as an independent observation to validate the ERD models. As root water-uptake is generally a non-fractionating process, tree $\mathrm{\delta}$^2^H~xylem~ reflects the signature of source water. Given a vertical gradient of $\mathrm{\delta}$^2^H in soil and groundwater, $\mathrm{\delta}$^2^H~xylem~ provides an index of rooting depth [@Dawson.1991]. We leveraged $\mathrm{\delta}$^2^H~xylem~ from BCI [@Meinzer:2001] for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species and vertically in soil and groundwater $\mathrm{\delta}$^2^H at natural abundance level (Fig. S3).

\
For comparison with modelled ERD, we removed six species from the @Meinzer:1999 dataset to account for the uncertainty in their water-sourcing depths. $\mathrm{\delta}$^2^H~xylem~ from leafless trees may not be linked with water sourced at the time of measurement and thus may not be comparable to species that had leaves. Leaflessness status of sampled trees is not recorded in @Meinzer:2001. We therefore removed five species that are typically leafless in March-April (Joseph Wright, personal observation), that is, the months of $\mathrm{\delta}$^2^H~xylem~ sampling. $\mathrm{\delta}^2H$ to depth relationship was particularly uncertain for $\mathrm{\delta}^2H$ > -40 ‰, so from the remaining dataset, we removed one species, *Guapira standleyana*, with $\mathrm{\delta}$^2^H~xylem~ of -28.9 ‰ $\\pm$ 3.7SE (see Fig. S3). For each model of ERD (Eq. \@ref(eq:M-chosen) and alternatives in SI), modeled species ERD was regressed against species $\mathrm{\delta}$^2^H~xylem~ for a maximum of six species.

@@

We introduce a novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil water potential profiles, and observed VPD (\@ref(eq:M-chosen), Figs. \@ref(fig:schematic), \@ref(fig:sp-erd)). Our predictions of ERD were consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Figs. \@ref(fig:erd-iso)). Our analyses suggest that co-occurring large canopy tree species with deeper ERD were associated with higher above-ground hydraulic efficiency, but lower safety (Fig. \@ref(fig:erd-hyd)). Nevertheless, evergreen but not deciduous species with deeper ERD showed significantly lower mortality rates than shallower ERD species (Fig. \@ref(fig:erd-mort-census-evg), S11). This ERD-mortality relationship was significant in five out of six census-intervals that had experienced one or more El-Niño events---over seven census-intervals studied in total (1982-2015). Quantifying extreme droughts in the whole soil column over 1990-2015 revealed that ERD explained significant levels of mortality in periods when extreme soil droughts prolonged the dry season water-stress (Fig. \@ref(fig:swp-droughts)). Species exposure to water-stress exponentially declined with deeper ERD (Fig. \@ref(fig:erd-time-evg)). Because extreme, sustained water-stress increases risk of mortality via hydraulic-failure and/or carbon starvation, deeper ERD may mitigate drought-induced mortality by limiting exposure to water-stress [@Rowland:2015; @McDowell:2018]. We thus demonstrate, for the first time, a link between species trade-offs in above- and below-ground hydraulic traits, drought exposure through extreme hydrological droughts quantified over the whole soil-column, and large ($\ge$ 10 cm DBH) tree mortality across several El-Niño events over 35 years.

@@

\
We found significant ERD-mortality relationships during extreme water-stress for evergreen but not deciduous species (Fig. \@ref(fig:erd-mort-census-evg), S11); consistent with the expectation that deciduous species can also avoid water-stress via leaf-drop. This is also consistent with the observation that species distribution along local (BCI) and regional (Panama) moisture gradients is correlated with $\mathrm{\Psi}$~tlp~ for evergreen but not deciduous species [@Kunert.2021]. Studies assessing coordination between leaf phenology, rooting depth and hydraulic traits simultaneously are almost absent for the tropics and warrant future consideration [@Oliviera.2021]. Our analyses for ERD-hydraulic traits relationships were limited to only seven species, but covered all species of different leaf phenologies (Fig. \@ref(fig:erd-hyd)), so here we assume that the trend in the ERD-hydraulic trait relationships holds true across all species of different leaf phenologies. We also found that the range of ERDs overlapped across leaf phenology, but evergreen species tend to have deeper ERDs (Fig. \@ref(fig:sp-erd)), consistent with observations by @Meinzer:1999 and elsewhere [@Smithmartin.2020; @Fan:2017; @Oliviera.2021]. Whether deciduous species also have more efficient and vulnerable hydraulics at BCI as is observed elsewhere remains to be studied [@Markesteijn:2010; @Markesteijn:2011; @Xu:2016; @Gleason.2016]. Leaf phenologies at BCI are numerous and complicated and warrant research.

@@

\
As in CT2018, with water-availability resolved for a 1-D column, we interpret our ERD estimates as revealing relative differences among species’ effective rooting depths rather than absolute depths.  To estimate absolute depths and topographic variation in ERD within or across species, future studies may use soil moisture dynamics from a distributed hydrological model [e.g. @Schwantes:2018]. We note that data availability on soil water retention curves ($\mathrm{\Psi}$~soil~ vs. VWC), hydraulic conductivity (K~soil~) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model. We recommend widespread and coordinated collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments.

@@

\
Figure \@ref(fig:erd-mort-census-evg): Mortality rate (mean $\pm$ 1SE in % yr^-1^) versus modeled effective rooting depth (ERD; mean $\pm$ 1SE in m) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen, canopy species found on Barro Colorado Island over seven census intervals (1981-2015). $R$^2^ and significance levels for linear model fits are given in panel insets. Model fits (blue lines) with confidence bands (gray area) are only shown for periods with significant fits at $\alpha$ = 0.05 or $\alpha$ = 0.1 (1982-85), identified by census-intervals with a superscript *.

\
Figure \@ref(fig:erd-time-evg): Modeled effective rooting depth (ERD; horizontal-axis) versus time spent beyond critical hydraulic threshold (vertical axis) by census interval (colored bars) for `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))` evergreen species included in the mortality analyses (Fig. \@ref(fig:erd-mort-census-evg)). Each bar represents the average time species of the same ERD spent beyond species-specific critical hydraulic thresholds in a given interval, that is, the proportion of days for which $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ was more negative than species $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~20,leaf~, and where $z$ is the soil depth matching species ERD. Standard error of the mean are shown over each bar when available. Note the squared y-axis scale.
